---
title: "Non spatial MOBSTER simulations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Non spatial MOBSTER simulations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


```{r packages, echo=TRUE, message=FALSE, warning=FALSE}
library(TEMULATOR) # package containing the simulator and datasets
library(ggplot2)
library(cowplot)  
theme_set(theme_cowplot())
library(gridExtra)
library(grid)
library(dplyr)
```


# Generation of the simulations

Below we define all the constant parameters. These are - unlike the seed, subclone birth rate (λ_sc) and time point of mutation (t_sc) - not varied. The birth rate of the ancestral clone is λ_a = 1.

```{r constant_parameters}
# Simulation parameters:
mutation_rate = 16   # mutation rate per cell doubling (m)
death_rate = 0.2     # death rate (μ)
end_time = 179782830 # number of reactions until the simulation in finished


# Sequencing parameters:
depth_model = 2   # over-dispersed beta-binomial distribution, dispersion parameter ρ=0.0
n_clonal = 500    # number of clonal mutations (N_clonal)
depth = 120       # 120x simulated coverage (C̅)
min_vaf = 0.05    # minimum VAF to accept a variant
```

For each combination of λ_sc and t_sc a total of 9 simulations were generated. Since calculating these data takes significant time a dataset containing a summary of all simulation ('mobster_summary_of_simulations') and a list of S3 objects containing the datasets used in the paper ('mobster_simulations') are included in this package.


The first dataset summaries all 1755 simulations:

```{r summary_of_precalculated_data}
data("mobster_summary_of_simulations", package="TEMULATOR")
colnames(mobster_summary_of_simulations)

t_sc = mobster_summary_of_simulations$subclone_start
b_sc = mobster_summary_of_simulations$subclone_birthrate

cat("t_sc: ", unique(t_sc), "\n")

cat("lambda_sc: ", unique(b_sc), "\n")

cat("Number of rows:", NROW(mobster_summary_of_simulations), "\n")

stopifnot(all(table(t_sc, b_sc) == 9))
```


The second dataset contains the simulations used in the paper:

```{r load_simulation_results, warning=FALSE}
data("mobster_simulations", package="TEMULATOR")

cat(length(mobster_simulations))

print(mobster_simulations[[1]])

plot(mobster_simulations[[1]])
```

These pre-calculated data are used throughout this vignette, but they could be recreated using the following code:

```{r how_to_simulate_results, eval=FALSE, include=TRUE}
for (i in seq_len(NROW(mobster_summary_of_simulations))) { 
  
  # the current variable parameters:
  sc_st = mobster_summary_of_simulations$subclone_start[i]
  sc_br = mobster_summary_of_simulations$subclone_birthrate[i]
  seed = mobster_summary_of_simulations$seed[i]
    
  
  simulation_result = 
    simulateTumour(
      birthrates = c(1.0, sc_br),
      deathrates = rep(death_rate, 2), 
      mutation_rates = rep(mutation_rate, 2),
      clone_start_times = c(0, sc_st),
      fathers = c(0, 0),                # CLONAL -> SC1 -> SC2, linear evolution 
      simulation_end_time = end_time, 
      seed = seed, 
      depth_model = depth_model, 
      min_vaf = min_vaf, 
      number_clonal_mutations = n_clonal,
      verbose=TRUE
    )
    
  
  # Check that the fraction of subclone cells are equal:
  x_sc = as.numeric(get_clone_frequency(simulation_result)["clone_2"])
  stopifnot(all.equal(mobster_summary_of_simulations$subclone_fraction[i], x_sc))
  
  
  # Check that the included dataset is identical:
  id = mobster_summary_of_simulations$assigned_id[i]
  if (!is.na(id)) { # one of the 150 selected simulations
    
    make_comparable = function(d) { 
      # drops two ids that can differ in the objects
      validate_temulator_result_object(d) 
      d$mutation_data$id = NULL 
      d$mutation_data$clone = NULL
      return(d)
    }
    
    included_data = make_comparable(mobster_simulations[[id]])
    calculated_data = make_comparable(simulation_result)
    stopifnot(all.equal(included_data, calculated_data))
  }
  
}
```




# Selection of suitable simulations


The plot below shows the fraction of all cells that subclone made up. For a given insertion time the subclone can either not grow to a significant size (e.g. λ_sc < 1.5 and t_sc = 2048) or grow so fast that it dominates the whole population (e.g. λ_sc > 2 and t_sc = 2048). In neither of these two cases a subclonal cluster would be detectable in the site frequency spectrum at a medium sequencing depth of C ≈ 120 (see figures below). Simulations where the subclone is at a very high or low frequency (red areas) are effectively neutral and from each 20 simulations were choosen at random

In order to create a reasonable tests with a detactable cluster in the site frequency spectrum 110 simulations where drawn from those where the subclone made up a fraction of 0.2-0.8 of all cells (green area in the plot blow). 


```{r fraction_subclonal_per_parameter_set, fig.width=6, fig.height=8}
poly_a = data.frame(x=c(-Inf,-Inf,Inf,Inf,-Inf), y=c(1.02,0.9,0.9,1.02,1.02))
poly_b = data.frame(x=c(-Inf,-Inf,Inf,Inf,-Inf), y=c(0.8,0.2,0.2,0.8,0.8))
poly_c = data.frame(x=c(-Inf,-Inf,Inf,Inf,-Inf), y=c(0.05,0,0,0.05,0.05))

mobster_summary_of_simulations %>% 
  filter(subclone_start>4) %>%
  mutate(subclone_start=factor(subclone_start, sort(unique(subclone_start)), ordered=1)) %>%
  mutate(subclone_birthrate_f=factor(subclone_birthrate, ordered=1)) %>%
  ggplot(aes(x=subclone_birthrate, y=subclone_fraction, group=subclone_birthrate_f)) + 
  geom_polygon(data=poly_a, aes(x=x, y=y), inherit.aes = FALSE, fill="red", alpha=0.5) + 
  geom_polygon(data=poly_b, aes(x=x, y=y), inherit.aes = FALSE, fill="green", alpha=0.2) + 
  geom_polygon(data=poly_c, aes(x=x, y=y), inherit.aes = FALSE, fill="red", alpha=0.5) + 
  geom_hline(yintercept = c(0.05, 0.2, 0.8, 0.9), linetype=3, color="gray10") +
  geom_boxplot() + 
  scale_x_continuous(breaks=c(1,1.5,2,2.5), labels=c("1",1.5,2, 2.5)) +
  facet_wrap(~subclone_start, labeller = as_labeller(function(x) paste0("t_sc = ", x)), ncol=3) + 
  xlab("Birthrate") + 
  ylab("Fraction of cells subclone")
```


```{r filter_results}
mobster_summary_of_simulations$group = 
  case_when(mobster_summary_of_simulations$subclone_fraction < 0.05 ~ "neutral_low_frac", 
            mobster_summary_of_simulations$subclone_fraction > 0.9 ~ "neutral_high_frac",
            between(mobster_summary_of_simulations$subclone_fraction, 0.2, 0.8)  ~ "detectable")


mobster_summary_of_simulations_subset = 
  mobster_summary_of_simulations %>% 
  filter(n_mutations_before_insertions >= 50) %>% 
  filter(!is.na(group)) %>% 
  filter(!is.na(assigned_id)) %>%
  arrange(subclone_fraction)


ids_per_group = 
  split(mobster_summary_of_simulations_subset$assigned_id,
        mobster_summary_of_simulations_subset$group)

sapply(ids_per_group, length)
```


## Plots group 1: Neutral with low subclone fraction

```{r}
plot_grid = function(object_list, title) {
  
  plots =
    mobster_simulations[ids_per_group[[set]]] %>% 
    lapply(plot, quite=TRUE)

  for (i in seq_along(plots)) {
    plots[[i]] = 
      plots[[i]] + 
      labs(title=substr(ids_per_group[[set]][i], 1, 10)) + 
      theme(plot.caption=element_text(hjust=0, size=6)) + 
      guides(fill=FALSE)
  } 

  top = textGrob(title,  gp=gpar(fontsize=25, fontface="bold"))
  grid.arrange(grobs=plots, top=top, ncol=3)
  
}

```


```{r plot_neutral_low_frac, fig.height=22, fig.width=12, out.width='90%', message=FALSE, warning=FALSE}
set = "neutral_low_frac"
title = "Effectively neutral - Low subclone frequency"
plot_grid(mobster_simulations[ids_per_group[[set]]], title)
```


## Plots group 2: Neutral with high subclone fraction

```{r neutral_high_frac, fig.height=22, fig.width=12, out.width='90%', message=FALSE, warning=FALSE}
set = "neutral_high_frac"
title = "Effectively neutral - High subclone frequency"
plot_grid(mobster_simulations[ids_per_group[[set]]], title)
```


## Plots group 3: Non-neutral with potentially detectable subclone

```{r detectable, fig.height=120, fig.width=12, out.width='90%', message=FALSE, warning=FALSE}
set = "detectable"
title = "Potentially detectable subclone"
plot_grid(mobster_simulations[ids_per_group[[set]]], title)
```

